---
title: "Lab 10 - Smoking during pregnacy"
subtitle: "Simulation based inference"
output: 
  html_document
link-citations: yes
---

```{r include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

In 2004, the state of North Carolina released a large data set containing information on births recorded in this state.
This data set is useful to researchers studying the relation between habits and practices of expectant mothers and the birth of their children.
We will work with a random sample of observations from this data set.

# Learning goals

-   Constructing confidence intervals
-   Conducting hypothesis tests
-   Interpreting confidence intervals and results of hypothesis tests in context of the data

# Getting started

Go to the course GitHub organization and locate your repo, clone it in RStudio and open the R Markdown document.
Knit the document to make sure it compiles without errors.

## Warm up

Let's warm up with some simple exercises.
Update the YAML of your R Markdown file with your information, knit, commit, and push your changes.
Make sure to commit with a meaningful commit message.
Then, go to your repo on GitHub and confirm that your changes are visible in your Rmd **and** md files.
If anything is missing, commit and push again.

## Packages

We'll use the **tidyverse** package for much of the data wrangling and visualization, the **tidymodels** package for inference, and the data lives in the **openintro** package.
These packages are already installed for you.
You can load them by running the following in your Console:

```{r}
library(tidyverse) 
library(tidymodels)
library(openintro)
library(skimr)
```

## Data

The data can be found in the **openintro** package, and it's called `ncbirths`.
Since the dataset is distributed with the package, we don't need to load it separately; it becomes available to us when we load the package.
You can find out more about the dataset by inspecting its documentation, which you can access by running `?ncbirths` in the Console or using the Help menu in RStudio to search for `ncbirths`.
You can also find this information [here](https://www.openintro.org/data/index.php?data=ncbirths).

# Set a seed!

In this lab we'll be generating random samples.
The last thing you want is those samples to change every time you knit your document.
So, you should set a seed.
There's an R chunk in your R Markdown file set aside for this.
Locate it and add a seed.


# Exercises

1.  What are the cases in this data set? How many cases are there in our sample?

The first step in the analysis of a new dataset is getting acquainted with the data.
Make summaries of the variables in your dataset, determine which variables are categorical and which are numerical.
For numerical variables, are there outliers?
If you aren't sure or want to take a closer look at the data, make a graph.

```{r explore-data}
skim(ncbirths)
```

**There are various ways to do this. One possibility is to use the skim function.**
**This easy function answers the number of observations and variables and the column type frequency.**
**We can also see if we have missing data and look quickly at means/sd**

```{r}
glimpse(ncbirths)
```

**The glimpse function can give us some information about which variable is which type.**

```{r}
ncbirths%>%
  ggplot(aes(x = fage)) + 
  geom_boxplot()
```

**And we can just repeat this for all of the different numerical variables: mage, weeks, visits, gained, weight**

## Baby weights

A 1995 study suggests that average weight of European American babies born in the US is 3,369 grams (7.43 pounds).[^lab-12-inference-smoking-1]
In this dataset we only have information on mother's race, so we will make the simplifying assumption that babies of European American mothers are also European American, i.e. `whitemom = "white"`.

We want to evaluate whether the average weight of European American babies has changed since 1995.

Our null hypothesis should state "there is nothing going on", i.e. no change since 1995: $H_0: \mu = 7.43~pounds$.

Our alternative hypothesis should reflect the research question, i.e. some change since 1995.
Since the research question doesn't state a direction for the change, we use a two sided alternative hypothesis: $H_A: \mu \ne 7.43~pounds$.

3.  Create a filtered data frame called `ncbirths_white` that contain data only from white mothers.
    Then, calculate the mean of the weights of their babies.
    
```{r}
ncbirths_white <- ncbirths%>%
  filter(whitemom == "white")

skim(ncbirths_white)
```
    

Let's discuss how this test would work.
Our goal is to simulate a null distribution of sample means that is centered at the null value of 7.43 pounds.
In order to do so, we

-   take a bootstrap sample of from the original sample,
-   calculate this bootstrap sample's mean,
-   repeat these two steps a large number of times to create a bootstrap distribution of means centered at the observed sample mean,
-   shift this distribution to be centered at the null value by subtracting / adding X to all bootstrap mean (X = difference between mean of bootstrap distribution and null value), and
-   calculate the p-value as the proportion of bootstrap samples that yielded a sample mean at least as extreme as the observed sample mean.

4.  Run the appropriate hypothesis test, visualize the null distribution, calculate the p-value, and interpret the results in context of the data and the hypothesis test.

**In this first set of code, let's just do a simple bootstrapping for a mean and calculate a 95% CI and visualize it. Then we'll add to this.**

```{r}
set.seed(1122)

boot_means <- ncbirths_white%>%
  specify(response = weight)%>%
  generate(reps = 1000, type = "bootstrap")%>%
  calculate(stat = "mean")

ggplot(data = boot_means, aes(x = stat)) + 
  geom_histogram()
```

**So the mean here looks different from that 7.43 in 1995...interesting**
**Now let's calculate the 95% CI**

```{r}
boot_means%>%
  summarize(
    lower = quantile(stat, 0.025),
    upper = quantile(stat, 0.975)
  )
```

**We are 95% confident that the mean of all White babies born in 2004 in NC is between 7.15 and 7.35 lbs**

**Since the null value is outside the CI, we would reject the null hypothesis in favor of the alternative. But let's test this.**

**So now let's do the hypothesis test for the difference**

```{r}
set.seed(1122)

boot_means_shifted <- ncbirths_white%>%
  specify(response = weight)%>%
  hypothesize(null = "point", mu = 7.43)%>%
  generate(reps = 1000, type = "bootstrap")%>%
  calculate(stat = "mean")

ggplot(data = boot_means_shifted, aes(x = stat)) + 
  geom_histogram() + 
  geom_vline(xintercept = 7.25, color = "red") + 
  geom_vline(xintercept = 7.43 + (7.43 - 7.25), color = "red")
```

```{r}
boot_means_shifted %>%
  filter(stat <= 7.25)%>%
  summarize(p_value = 2 * (n() / 1000))
```

*Since p-value is less than the significance level, we reject the null hypothesis. The data provide convincing evidence that the average weight of White babies in NC in 2004 is different from 1995 weights**


## Baby weight vs. smoking

Consider the possible relationship between a mother's smoking habit and the weight of her baby.
Plotting the data is a useful first step because it helps us quickly visualize trends, identify strong associations, and develop research questions.

5.  Make side-by-side boxplots displaying the relationship between `habit` and `weight`.
    What does the plot highlight about the relationship between these two variables?
    
```{r}
ncbirths%>%
  ggplot(aes(x = weight, y = habit)) + 
  geom_boxplot()
```
    
    **It looks like the weight of nonsmokers is slightly higher, but also that there is a lot more variability.**

6.  Before moving forward, save a version of the dataset omitting observations where there are NAs for `habit`.
    You can call this version `ncbirths_habitgiven`.
    
```{r}
ncbirths_habitgiven <-  ncbirths%>%
  filter(!is.na(habit))
```
    

The box plots show how the medians of the two distributions compare, but we can also compare the means of the distributions using the following to first group the data by the `habit` variable, and then calculate the mean `weight` in these groups using.

```{r habit-means, eval=FALSE}
ncbirths_habitgiven %>%
  group_by(habit) %>%
  summarise(mean_weight = mean(weight))
```

There is an observed difference, but is this difference statistically significant?
In order to answer this question we will conduct a hypothesis test .

7.  Run the appropriate hypothesis test, calculate the p-value, and interpret the results in context of the data and the hypothesis test.

**What type of test is this? Independent samples t-test**
**In this code a few things change:**
**I need to specify the explanatory variable and I change the type of simulation to permute.**
**I am also calculating a t-statistic and I need to specify the order of subtraction.**
**This order is saying to find a mean difference between nonsmokers and smokers**

```{r}
set.seed(1122)

diff_means <- ncbirths_habitgiven%>%
  specify(response = weight, explanatory = habit)%>%
  hypothesize(null = "independence")%>%
  generate(reps = 1000, type = "permute")%>%
  calculate(stat = "t", order = c("nonsmoker", "smoker"))
```

**And here is a different way to calculate the p-value**
**The observed stat comes from our raw calculations on the sample data**

```{r}
diff_means%>%
  get_p_value(obs_stat = 0.31, direction = "greater")
```


***Oooohhhh that difference is not significant. But let's do the confidence interval anyways***

8. Construct a 95% confidence interval for the difference between the average weights of babies born to smoking and non-smoking mothers.

```{r}
diff_means%>%
  summarize(
    lower = quantile(stat, 0.025),
    upper = quantile(stat, 0.975)
  )
```

**We are 95% confident that the difference between means ranges from -2.11 and 1.86. It contains 0 which supports the p-value being non-sig.**


## Baby weight vs. mother's age

In this portion of the analysis we focus on two variables.
The first one is `maturemom`.

9. First, a non-inference task: Determine the age cutoff for younger and mature mothers. Use a method of your choice, and explain how your method works.

***Hmmmm, so would we do the median? Would we use a medical cutoff for geriatric pregnancies? 

```{r}
ncbirths%>%
  skim()
```

**The mean age is 27. So could we cut that way? Or would we maybe do the 75th percentile?**


The other variable of interest is `lowbirthweight`.

10. Conduct a hypothesis test evaluating whether the proportion of low birth weight babies is higher for mature mothers. State the hypotheses, run the test and calculate the p-value, and state your conclusion in context of the research question. Use $\alpha = 0.05$. If you find a significant difference, construct a confidence interval, at the equivalent level to the hypothesis test, for the difference between the proportions of low birth weight babies between mature and younger mothers, and interpret this interval in context of the data.



**Very similar to our previous example. But this time I am dealing with proportions. So I need to tell it what a "success" is.**
**This is the outcome we care about**
**Everything else looks similar to our t-test, but this time I specify "diff in props"**

```{r}
set.seed(1122)

mature_diff <- ncbirths%>%
  specify(response = lowbirthweight, explanatory = mature, success = "low")%>%
  hypothesize(null = "independence")%>%
  generate(reps = 1000, type = "permute")%>%
  calculate(stat = "diff in props", order = c("mature mom", "younger mom"))
  
```


**Calculate the p-value. But first we really need to know the proportions for each group to get the observed value to test against.** 

```{r}
prop_mature <- ncbirths%>%
  group_by(mature, lowbirthweight)%>%
  summarize(cnt = n())%>%
  mutate(freq = round(cnt / sum(cnt), 3))

```

**Mature moms have .135 lowbirthweight and young have .107**

```{r}
mature_diff %>%
  filter(stat >= 0.028) %>%
  summarize(p_value = n()/nrow(mature_diff))
```

**Or we can try the automatic way**

```{r}
mature_diff%>%
  get_p_value(obs_stat = 0.028, direction = "greater")
```

**We get the same answer**


[^lab-12-inference-smoking-1]: Wen, Shi Wu, Michael S. Kramer, and Robert H. Usher.
    "Comparison of birth weight distributions between Chinese and Caucasian infants." American Journal of Epidemiology 141.12 (1995): 1177-1187.
